A Mechanised Library for Generic (Co)Programming

### 2024-03-05 paper/paper.tex
Algebras and Coalgebras

\subsection{Algebras}
\label{sec:algebras}
The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

A fold consumes elements of an inductive data type. These can be thought as
least ``sets'' generated by some constructors. For example, the type of lists
over a type \coq{A}, namely \coq{List A} is the ``least set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
Together, \coq{Empty} and \coq{Cons} form an \emph{algebra} of
type \coq{In : unit + A * List A -> List A}. In particular, this algebra is an
isomorphism with the obvious inverse \coq{InOp : List A -> unit + A * List A}
returning \coq{tt} is the list is empty and the head and tail of the list
otherwise. This type of algebra is called \emph{initial algebra}.  This is
because for for any other algebra \coq{alg : unit + A * List B -> B} there
exists a unique map which we call \coq{cata alg : List A -> B} which destructs a
list into its head and tail using a
destructor \coq{InOp : List A -> unit + A * List A}, applies itself recursively,
composes together the results of the recursion using the algebra \coq{alg}.

To generalise recursion schemes we abstract the shape of the data using functors.
Assuming a base functor \coq{F} a catamorphism is then the unique
solution to the equation
\begin{minted}{coq}
cata alg = alg . fmap (cata alg) . InOp
\end{minted}
which does precisely what we described in the previous paragraph.

@@

Given a type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the
carrier `\coq{A}` is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
it is called the \emph{initial algebra} and denotes the inductive data type shaped by the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

@@

We define \emph{catamorphisms} by recursively traversing their structure of the
input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}

@@

Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
\begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
\caption{Terminal Coalgebras}\label{fig:term_coalg}
\end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

>>>>

\subsection{Algebras}
\label{sec:algebras}
An algebra is a set of abstract operations on a type \coq{X}. For example,
give such a type \coq{X} we can define the list operations on \coq{X} as
\coq{Empty : unit -> X} and \coq{Cons : A * X -> X}.

We can generalise this using containers. Given a type `\coq{A}` and a
container \coq{F} an algebra for \coq{F} over the carrier `\coq{A}` is a map of
the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

The type of lists over a type \coq{A}, namely \coq{List A} is the ``least set''
generated by the constructors \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.  Together, \coq{Empty} and \coq{Cons}
form an \emph{algebra} of type \coq{In : unit + A * List A -> List A}.

When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
it is called the \emph{initial algebra} and denotes the inductive data type
shaped by the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
In particular, this algebra is an isomorphism with the obvious
inverse \coq{InOp : List A -> unit + A * List A} returning \coq{tt} is the list
is empty and the head and tail of the list otherwise. This type of algebra is
called \emph{initial algebra}.  This is because for for any other
algebra \coq{alg : unit + A * List B -> B} there exists a unique map which we
call \coq{cata alg : List A -> B} which destructs a list into its head and tail
using a destructor \coq{InOp : List A -> unit + A * List A}, applies itself
recursively, composes together the results of the recursion using the
algebra \coq{alg}.

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

A fold consumes elements of an inductive data type. These can be thought as
least ``sets'' generated by some constructors.
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.

@@

@@

We define \emph{catamorphisms} by recursively traversing their structure of the
input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}
\subsection{Coalgebras}
Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
\begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
\caption{Terminal Coalgebras}\label{fig:term_coalg}
\end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras.

\subsubsection{On Equality}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

@@

In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
